---
title: "Wine Quality Challenge"
author: "Luca Marchiori"
date: "2024-03-27"
output:
  html_document: default
  pdf_document: default
---
# Wine Quality Challenge
## Instructions
The inputs include objective tests (e.g. PH values) and the output is based on sensory data (median of at least 3 evaluations made by wine experts). Each expert graded the wine quality between 0 (very bad) and 10 (very excellent).

Submissions are evaluated on Root-Mean-Squared-Error (RMSE) between the predicted value and the observed quality. 

RMSE = sqrt(mean((y – haty)^2)) 

## Dataset colums
1. fixed acidity
2. volatile acidity
3. citric acid
4. residual sugar
5. chlorides
6. free sulfur dioxide
7. total sulfur dioxide
8. density
9. pH
10. sulphates
11. alcohol
12. quality (score between 0 and 10)

## Init
```{r}
# Load libraries
library(caret)
library(ggplot2)
library(corrplot)
library(ggbiplot)
library(rpart)
library(rpart.plot)
library(randomForest)
library(leaps)
#Clear the workspace
rm(list=ls())
```

## Data import
```{r}
# Load the training data
train <- read.csv("wineq_train.csv", stringsAsFactors=F)
# Load the test data
test <- read.csv("wineq_validation.csv", , stringsAsFactors=F)
```
## Helpers
### Minimal Squared Error
MSE (Mean Squared Error) is a common metric for evaluating the performance of regression models, by comparing the observed values `y` with the predicted values `yhat` generated by the provided `model`.
```{r}
MSE <- function(y, model){
  yhat = predict(model) # Estimated y computed on the basis of the features
  mean((y-yhat)^2)
}

RMSE <- function(y, model){
  sqrt(MSE(y,model))
}
```
## Exploratory Data Analysis
Initial investigation and analysis of the dataset to understand its key characteristics, identify data quality issues and understand relationships between variables.
```{r}
# Compactly Display the Structure of the dataset
str(train)

# Take an initial view of the dataset
summary(train)
```
All the features are numeric, no qualitative data is present. The quality is the target variable and since it can be considered as a continuous variable, we can use regression models to predict it.

### Outliers detection
```{r}
# Boxplot of the features
par(mfrow=c(3,4))
for (i in 1:12) {
  boxplot(train[,i], main = colnames(train)[i])
}

# Plot feature values to find outliers
plot(train$residual.sugar, main = "Residual Sugar", pch = 19, cex = 0.5)
plot(train$density, main = "Density", pch = 19, cex = 0.5)
plot(train$chlorides, main = "Chlorides", pch = 19, cex = 0.5)
plot(train$free.sulfur.dioxide, main = "Free Sulfur Dioxide", pch = 19, cex = 0.5)
plot(train$total.sulfur.dioxide, main = "Total Sulfur Dioxide", pch = 19, cex = 0.5)
plot(train$volatile.acidity, main = "Volatile Acidity", pch = 19, cex = 0.5)
plot(train$fixed.acidity, main = "Fixed Acidity", pch = 19, cex = 0.5)
plot(train$pH, main = "pH", pch = 19, cex = 0.5)
plot(train$sulphates, main = "Sulphates", pch = 19, cex = 0.5)
plot(train$alcohol, main = "Alcohol", pch = 19, cex = 0.5)
plot(train$citric.acid, main = "Citric Acid", pch = 19, cex = 0.5)
plot(train$quality, main = "Quality", pch = 19, cex = 0.5)
```
Outliers listed below are removed from the dataset because they are likely to be errors in the data since they are very far from the other values.
```{r}
# Remove dataset feature if residual sugar is higher than 40
train <- train[train$residual.sugar < 40,]
# Remove dataset feture if density is higher than 1.005
train <- train[train$density < 1.005,]
# Remove dataset feature if free sulfur dioxide is higher than 200
train <- train[train$free.sulfur.dioxide < 200,]
# Remove dataset feature if total sulfur dioxide is higher than 350
train <- train[train$total.sulfur.dioxide < 350,]
# Remove dataset feature if fixed acidity is higher than 15
train <- train[train$fixed.acidity < 15,]
```


### Scatterplots
With the following scatterplots, we can see the relationship between the quality and the other features. Each level of quality is represented by a different color.
```{r}
# Scatterplot matrix, with different colors for each quality level
pairs(train, col = train$quality, upper.panel = NULL, pch = 19, cex = 0.5)

# Single scatterplots of features against quality
# plot(train$alcohol, train$quality, col = train$quality, pch = 19, cex = 0.5)
# plot(train$total.sulfur.dioxide, train$quality, col = train$quality, pch = 19, cex = 0.5)
# plot(train$free.sulfur.dioxide, train$quality, col = train$quality, pch = 19, cex = 0.5)
# plot(train$volatile.acidity, train$quality, col = train$quality, pch = 19, cex = 0.5)
# plot(train$fixed.acidity, train$quality, col = train$quality, pch = 19, cex = 0.5)
# plot(train$pH, train$quality, col = train$quality, pch = 19, cex = 0.5)
# plot(train$density, train$quality, col = train$quality, pch = 19, cex = 0.5)
# plot(train$sulphates, train$quality, col = train$quality, pch = 19, cex = 0.5)
# plot(train$chlorides, train$quality, col = train$quality, pch = 19, cex = 0.5)
# plot(train$residual.sugar, train$quality, col = train$quality, pch = 19, cex = 0.5)
# plot(train$citric.acid, train$quality, col = train$quality, pch = 19, cex = 0.5)

```
### Principal Component Analysis
Here, PCA is used to identify the most important features in the dataset in order to better understand the data. Here it is only used for visualization purposes, as trees and random forests are better suited for feature selection in this case. The biplot shows the relationship between the features and the quality of the wine. 
By default, the prcomp() function, with the option "scale = true", centers and scales the data, so we don't need to do it manually.
```{r}
# Principal Component Analysis
pca <- prcomp(train[,1:12], scale = TRUE)
summary(pca)

# Biplot with GGplot2
ggbiplot::ggbiplot(pca, obs.scale = 1, var.scale = 1, groups = train$quality, ellipse = TRUE, circle = TRUE, varname.size = 3, varname.adjust = 3, varname.color = "orange", varname.face = "bold") +
  theme_minimal() +
  theme(legend.title = element_text(size = 10), legend.text = element_text(size = 10)) +
  ggtitle("PCA Biplot")
```

### Correlation heatmap
The correlation heatmap shows the correlation between the features. The correlation matrix is calculated using the cor() function, and the heatmap is plotted using the corrplot() function. In this dataset, alcohol, density, chlorides, and volatile acidity are the most correlated features with quality.
```{r}
corrplot(cor(train[,1:12]), method = "number", tl.col = "black", tl.cex = 0.7, number.cex = 0.7)
```

## Linear Model

### Model training

The wine quality challenge provides a test set of 1200 observations to evaluate the model. Instead of use it as a validation set, we will use the cross-validation technique to evaluate the model. The choice falls on a LOOCV (Leave-One-Out Cross-Validation) because the dataset is not too large, because we are evaluating a linear model (for wich LOOCV is optimezed), and because it is a good way to avoid overfitting.



First of all, a linear model is trained on all the features of the dataset. The summary of the model is printed to understand the imporntance of each feature. The quality of the test set is predicted and saved in the file for the datachallenge submission. The RMSE of the model is calculated and printed.
```{r}
# Train the linear model on all the features of the dataset
lmfit <- lm(quality ~ ., data = train)

# Get the summary of the model
summary(lmfit)

# Predict the quality of the test set
yhat = predict(lmfit, newdata=test)
write.table(file="./DatachallengeSubmissions/fitLinearModelAllFeatures.txt", yhat, row.names = FALSE, col.names = FALSE)

# Calculate the RMSE of the model
rmse = RMSE(y=train$quality, model=lmfit)
rmse # 0.7482341
```

Now we use the stepwise selection to select the most important features for the linear model. The model is trained with the selected features and the RMSE is calculated again.

```{r}
# Define the null model (no predictors)
nullModel <- lm(quality ~ 1, data = train)

# Define the full model (all predictors)
fullModel <- lm(quality ~ ., data = train)

# Perform forward stepwise selection
FwStepSel <- step(nullModel, direction = "forward", scope = list(lower = nullModel, upper = fullModel))

# Get the selected features
selectedFormula <- formula(FwStepSel)

# Fit the linear model with the selected features
fit <- lm(selectedFormula, data = train)

# Predict the quality of the test set
yhat = predict(fit, newdata=test)

# Save the predictions to a file
write.table(file="./DatachallengeSubmissions/fitLinearModelSubsetFeatures.txt", yhat, row.names = FALSE, col.names = FALSE)

# Calculate the RMSE of the model
rmse = RMSE(y=train$quality, model=fit)
rmse # 0.7483839
```

The RMSE of the model with all features is lower than the one with the selected features but this does not mean that the model with all features is better (since it can lead to overfitting the training data). In fact, the model with the selected features is better when loaded to the DataChallenge platform.

## Tree Models
In this section, decision trees and random forests are used to predict the quality of the wine. The models are trained with all the features and with the selected features chosen by the tree model results.

### Decision trees
A complete decision tree is trained on all the features of the dataset. The importance of each feature is calculated and plotted.
```{r}
# Fit a complete decision tree
dTreefit <- rpart(quality ~ ., data = train, method = "anova", cp = 0.005)

# Plot the decision tree
rpart.plot(dTreefit, type = 4, extra = 101, under = TRUE, fallen.leaves = TRUE, cex = 0.5, tweak = 1.2)

# Get the importance of each feature
feature_importance <- data.frame(Feature = names(dTreefit$variable.importance), Importance = dTreefit$variable.importance)

# Plot feature importance
barplot(feature_importance$Importance, names.arg = feature_importance$Feature, las = 2,main = "Feature Importance")

# Predict the quality of the test set
yhat = predict(dTreefit, newdata=test)
write.table(file="./DatachallengeSubmissions/fitDecisionTreeAllFeatures.txt", yhat, row.names = FALSE, col.names = FALSE)

# Calculate the RMSE of the model
rmse = RMSE(y=train$quality, model=dTreefit)
rmse # 0.7250806
```
From the decision trees importance results, we can exclude the features citric.acid, pH, fixed.acidity, and sulphates. Now we train a decision tree with the selected features.
```{r}
# Fit a decision tree on a fewer features
dTreefit <- rpart(quality ~ . -citric.acid - pH - fixed.acidity - sulphates, data=train)
rpart.plot(dTreefit)

# Predict the quality of the test set
yhat = predict(dTreefit, newdata=test)
write.table(file="./DatachallengeSubmissions/fitDecisionTreeSubsetFeatures.txt", yhat, row.names = FALSE, col.names = FALSE)

# Calculate the RMSE of the model
rmse = RMSE(y=train$quality, model=dTreefit)
rmse # 0.7488062
```

### Random forests
As an alternative to decision trees, random forests are used to predict the quality of the wine. The model is trained with all the features and with the selected features chosen by the random forest results.
```{r}
# Train the random forest model with all features
rForestfit <- randomForest(quality ~ ., train)

# Get the importance of each feature and plot it
importance <- importance(rForestfit)
varImpPlot(rForestfit)

# Predict the quality of the test set with all features
yhat = predict(rForestfit, newdata=test)

# Save the predictions to a file
write.table(file="./DatachallengeSubmissions/fitRandomForestAllFeatures.txt", yhat, row.names = FALSE, col.names = FALSE)

# Calculate the RMSE of the model
rmse = RMSE(y=train$quality, model=rForestfit)
rmse # 0.6031999
# DataChallenge submission: 0,6265 
```

Now we train a random forest model with the selected features chosen by the random forest results.

```{r}
# Train the random forest model with selected features
#rForestfit <- randomForest(quality ~ . - sulphates - fixed.acidity - citric.acid - pH, train)
rForestfit <- randomForest(quality ~ . - sulphates - fixed.acidity, train)

# Predict the quality of the test set with selected features
yhat = predict(rForestfit, newdata=test)

# Save the predictions to a file
write.table(file="./DatachallengeSubmissions/fitRandomForestSubsetFeatures.txt", yhat, row.names = FALSE, col.names = FALSE)

# Calculate the RMSE of the model
rmse = RMSE(y=train$quality, model=rForestfit)
rmse # 0.6084313
# DataChallenge submission: 0,6424
```



